---
title: "Analyses - CCFRP Angler Survey Publication"
author: "Erin Johnston"
date: "10/29/2023"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F)

```

Analyses after data processing steps:

#### Research Question 1

The dataset `opinion_change_dat` contains vars on 256 CCFRP volunteer angler respondents. The outcome variable is **qual_delta**, categories on positive change, negative change, and no change. The predictor vars are **region**, North/Central/South, and **num_trips** which asks categorically the number of CCFRP respondents had participtaed on with categories as follows:

* 1 = 1
* 2 = 2-5
* 3 = 6-10
* 4 = 11-19
* 5 = 20+

```{r multinomial}

opinion_change_dat$qual_delta2 <- relevel(opinion_change_dat$qual_delta, ref = "No Change")

model_op_change <- multinom(qual_delta2 ~ region + num_trips , data = opinion_change_dat)
summary(model_op_change)

z <- summary(model_op_change)$coefficients/summary(model_op_change)$standard.errors
z

p <- (1 - pnorm(abs(z), 0, 1)) * 2
p

```

#### Research Questions 2

Perceptions vs Ecological data
This will first be analyzed for **statewide angler responses**

```{r}

perceived_mpa <- dat_processed %>%
  select(id, q3, region, q15, q16, q17, q18) %>%
  filter(q15 == 1) %>%
  mutate(q16 = as.factor(q16),
         q17 = as.factor(q17),
         q18 = as.factor(q18))

perception_dat <- perceived_mpa %>%
  filter(!q16 == 3, !q17 == 3, !q18 == 3) %>%
  select(region, q16, q17, q18)

write_csv(perception_dat, here("Data", "perception_dat.csv"))

### abundance
## state
high_fish_state <- perceived_mpa %>%
  filter(!q16 == 3) %>%
  count(q16)

file_path <- "https://www.sthda.com/sthda/RDoc/data/housetasks.txt"
housetasks <- read.delim(file_path, row.names = 1)
dt <- as.table(as.matrix(housetasks))
balloonplot(t(dt), main ="housetasks", xlab ="", ylab="",
            label = FALSE, show.margins = FALSE)
wut <- chisq.test(housetasks)
wut$observed
wut$expected
corrplot(wut$residuals, is.cor = FALSE)

high_fish_statetest <- perceived_mpa %>%
  filter(!q16 == 3) %>%
  group_by(region) %>%
  count(q16) %>%
  mutate(q16 = case_when(
    q16 == "1" ~ "MPA",
    q16 == "2" ~ "REF",
    q16 == "4" ~ "No Difference"
  )) %>%
  pivot_wider(names_from = q16,
              values_from = n)

forbplot <- high_fish_statetest %>% remove_rownames %>% column_to_rownames(var="region")

dtest <- as.table(as.matrix(forbplot))
balloonplot(t(dtest), main ="test test", xlab ="", ylab="",
            label = FALSE, show.margins = FALSE)

lab_chisq <- chisq.test(forbplot[3,])
lab_chisq <- chisq.test(forbplot)
lab_chisq$observed
lab_chisq$expected

library(corrplot)
corrplot(lab_chisq$residuals, is.cor = FALSE)

## central CA
high_fish_c <- perceived_mpa %>%
  filter(!q16 == 3, region == "Central") %>%
  count(q16)

## southern CA
high_fish_s <- perceived_mpa %>%
  filter(!q16 == 3, region == "Southern") %>%
  count(q16)

## northern CA
high_fish_n <- perceived_mpa %>%
  filter(!q16 == 3, region == "Northern") %>%
  count(q16)

### diversity
## state
high_diversity_state <- perceived_mpa %>%
  filter(!q17 == 3) %>%
  count(q17)

## central CA
high_diversity_c <- perceived_mpa %>%
  filter(!q17 == 3, region == "Central") %>%
  count(q17) %>%
  add_row(q17 = "2", n = 0)

## southern CA
high_diversity_s <- perceived_mpa %>%
  filter(!q17 == 3, region == "Southern") %>%
  count(q17)

## northern CA
high_diversity_n <- perceived_mpa %>%
  filter(!q17 == 3, region == "Northern") %>%
  count(q17) 

### size
## state
high_size_state <- perceived_mpa %>%
  filter(!q18 == 3) %>%
  count(q18)

## central CA
high_size_c <- perceived_mpa %>%
  filter(!q18 == 3, region == "Central") %>%
  count(q18)

## southern CA
high_size_s <- perceived_mpa %>%
  filter(!q18 == 3, region == "Southern") %>%
  count(q18)

## nortern CA
high_size_n <- perceived_mpa %>%
  filter(!q18 == 3, region == "Northern") %>%
  count(q18)

```

```{r chi square tests}

exp <- rep(1/3, 3)

## statewide chi square tests of equal frequency
chisq.test(high_fish_state$n, p = exp)
chisq.test(high_diversity_state$n, p = exp)
chisq.test(high_size_state$n, p = exp)

## central coast chi square tests of equal frequency
chisq.test(high_fish_c$n, p = exp)
chisq.test(high_diversity_c$n, p = exp)
chisq.test(high_size_c$n, p = exp)

## southern coast chi square tests of equal frequency
chisq.test(high_fish_s$n, p = exp)
chisq.test(high_diversity_s$n, p = exp)
chisq.test(high_size_s$n, p = exp)

## norhtern coast chi square tests of equal frequency
chisq.test(high_fish_n$n, p = exp)
chisq.test(high_diversity_n$n, p = exp)
chisq.test(high_size_n$n, p = exp)

```


#### Research Question 3

```{r}

### why volunteer with CCFRP
why_volunteer <- dat_processed %>%
  select(id, q3, region, q19) %>%
  mutate(q19 = as.factor(q19))

why_volunteer <- why_volunteer %>%
  count(q19) %>%
  drop_na() %>%
  mutate(perc = n/sum(n))

exp_q19 <- rep(1/5, 5)

chisq.test(why_volunteer$n, p = exp_q19)

### do anglers tell their peers about CCFRP
peer_comm <- dat_processed %>%
  select(id, q3, region, q8) %>%
  mutate(q8 = as.factor(q8))

peer_comm <- peer_comm %>%
  count(q8) %>%
  drop_na() s

exp_q8 <- rep(1/2, 2)

chisq.test(peer_comm$n, p = exp_q8)


```

